---
title: Are old apartments cheaper? Part 1
author: ''
date: '2020-09-26'
slug: are-old-apartments-cheaper-part-1
categories: []
tags: []
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
```

```{r libs}
library(tidyverse)
library(gt)
library(here)
```




```{r data, cache = T}




url <- "https://raw.githubusercontent.com/Codecademy/datasets/master/streeteasy/streeteasy.csv"
d_raw <- read_csv(url)

# choosing most relevant features
keep_cols <- 
  
   c( "rent",
     "bedrooms",
     "bathrooms",
     "size_sqft",
     "neighborhood",
     "building_age_yrs",
     "borough",
     "submarket")

df <- d_raw %>% select(one_of(keep_cols))
df <- rename(df, age = building_age_yrs)
df <- mutate(df, borough = if_else(neighborhood == "Long Island City", "Queens", borough))

```


# Introduction

There are a lot of constraints to the housing supply in New york. Housing affordability is a major concern and national issue. A major theme in the discussion of how we can provide more affordable housing is that, contrary to Econ 101 logic, increases in the housing supply increase the median housing price/rent. one thing that probably isn't mentioned enough is that housing tends to get cheaper. in the economics literature this effect is called filtering effects-- as high-end apartments and new homes age, they tend to become more affordable.  in this post i will:

- introduce a dataset of 5000 NYC apartments for rent from the Real Estate platfor Streeteasy

- enhance this dataset with census data in order to make comparisons across neighborhood income levels

- discuss how this data will be used to identify filtering effects in the NYC rental market

## Filtering in the housing market

Filtering is an important idea in housing policy, expecially when it comes to maintaining a healthy supply of affordable housing. Filtering effects are very intuitive: as the housing stock becomes older, it depreciates, deteriorates and falls out of fashion. The idea that new, more expensive housing will *filter* down to become more affordable housing. If there are strong filtering effects in a housing market then, if a policy goal is to provide more affordable housing, increasing the current supply of higher end housing will probably increase the supply of affordable housing years in the future. 

A housing supply and affordability has become an important topic in the past decades, many efforts have gone to identifying these effects. (Coulson and Bond, 1990) take a hedonic-regression approach with a linear model and dont find strong evidence of filtering at all[^3]. In a future post ill explore the limitations of using linear methods for this problem. Economist Eric Rosenthal constructs an impressive panel dataset and finds evidence for both non-linear and heterogeneous filtering effects.[^4]




### The Streeteasy dataset

The real estate platform [Streeteasy](https://streeteasy.com/) released a dataset of 5,000 apartments for rent on their website. This dataset is accessible on [github](https://raw.githubusercontent.com/Codecademy/datasets/master/streeteasy/streeteasy.csv) and [Code Academy](https://www.codecademy.com/articles/streeteasy-dataset) does a good job of describing it. The most importants features for the sake of this project are:

`rent - price of rent ($)`</br>
`bedrooms - number of bedrooms`</br>
`size_sqft - size (ft²)` and </br>
`neighborhood - neighborhood (ex: Greenpoint)`

What is unique about this dataset is that the neighborhood is highly-granular. This allows me to assess neighborhood effects. For instance in a lot of other housing-unit level datasets, the granularity of the observation on the level of a Metropolitan Statistical Area (ex: New York MSA). This dataset allows me to compare similar units in Brownsville, Brooklyn to those in  Midtown, Manhattan.

#### Sampling

An important drawback of this dataset is not only that it only includes NYC apartments, but it is clearly not a random-sample of NYC apartments. This is evidenced by the fact that the sample doesn't include any units in the Bronx. My hunch is that Streeteasy apartments are fundamnetally different than a randomly selected NYC apartment. Exploring these *selection effects* is beyond the scope of this project but need to be kept in mind throughout.



```{r plot_fns}
library(patchwork)
library(glue)

scale_y_continuous <- function(...){
  ggplot2::scale_y_continuous(expand = c(0,0))
}
segment <- function(...){
  ggplot2::annotate(...,
                    geom  = "segment",
                    size  = 1.2,
                    alpha = .7,
                    arrow = arrow(length = unit(3, "mm"), ends = "first"))
}

geom_vline_t <- function(...){
  ggplot2::geom_vline(...,
                      size = 1.2,
                      alpha = .7)
}
```

## The long tail of rentals

New York City is full of contrasts. The high-rise luxury rentals overlooking Park Avenue are just blocks away from the crowded and dilapidated housing projects of Harlem. These contrasts are reflected (to a lesser extent) in the distribution of rental prices in the Streeteasy sample and its long right tail. These distributions are common in economic and social-science data.[^6]

```{r  log_rent, fig.align = "center", fig.height=3, fig.width=10}

theme_set(theme_minimal()+
            theme(legend.title = element_text(size = 5)))


foo <- 
df %>% 
  ggplot()+
  geom_histogram(aes(rent), fill = "grey80", color = "grey20")+
  theme_minimal()+
  labs(title = "rent distribution")

bar <- 

df %>% 
ggplot()+
  geom_histogram(aes(rent), color = "grey20", fill = "grey80")+
  scale_x_log10()+
  labs(title = ("log(rent) distribution"))+
  theme_minimal()

dsgn <- "
AAAA#BBBB
"

foo+bar+plot_layout(design = dsgn)


```



It is said that wealth creates wealth. This is why, according to oxfam the 1% richest people in the world control 44% of the wealth[^2]. This is also why so much economic data follow power law distributions-- or distributions that change in magnitude rather than absolutes. As rent is a function of income (wealthy people rent more expensive apartments) I'd expect rental prices to follow a power law distribution. Ultimately, I want to model the effects of age on rental price with a regression model. A key assumption of regression techniques are that the dependent variable in a linear combination of *additive* independent variables. However, without tranforming my data, my model will take on some functional form:

$$rent_i = e^{\beta_{1}x_{i1} + \beta_{2}x_{i2}...\beta_{k}x_{ik} + \epsilon_i}= e^{\beta_{1}x_{i1}}\times{e^{\beta_{2}x_{i2}}}\times{...}\times{e^{\beta_{k}x_{ik}}}\times\epsilon_i\:\:(i=1,2,...,n)$$


where $rent_i$ is the rent for the $i_{th}$ unit and $\beta_j$ is the $j_{th}$ predictor variable.

To fix this all we have to do is take the natural log of both sides of our equations:

$$log(rent_i) = \beta_1x_{i1} + \beta_2x_{i2} + \beta_3x_{i3} + \epsilon_i$$

and voila! our multiplicative model is an additive model. With our dependent variable on the log scale-- we can view the log of rent as the magnitude of rent.[^1] 


#### Whats a 1 standard deviation increase in rent?

```{r, include=T, echo=T}
# create ln(rent)
df$rent_log <- log(df$rent, base = exp(1))

# standard deviation of log(rent)
sigma_rent_log <- sd(df$rent_log)

# exponentiate sigma of log rent
(exp_sigma <- round(exp(sigma_rent_log), 2))

```

So, $e^\sigma=1.67$ which means a standard deviation increase in rent is a 67% inrease in the price of rent. 





```{r panel1, fig.align = "center", fig.height=3, fig.width=4}


age_dens <-
df %>% 
  ggplot()+
  geom_histogram(aes(age), bins = 60, color = "black", fill = "grey80")+
  geom_vline_t(xintercept = median(df$age))+
  
  segment(x = median(df$age)+2,
           xend = median(df$age)+30,
           y = 300,
           yend = 360)+
  annotate(geom = "text",
           x = 100,
           y = 385,
           label = "Median Age= 44",
           hjust = .5)+
  labs(y = "# of Units",
       x = "Age of building (yrs)",
       title = str_wrap("Multi-modal distribution of building age", 30))
dsgn <- "
AAA#BBB
"




```


### Bucketing units into *High*, *Low* and *Middle* income PUMA's

 

```{r bucketing, cache=T}

# census income data

library(tidycensus)
puma_incomes <- 
  get_acs(geography     = "public use microdata area",
          variables     = "B07011_001",
          geometry      = F,
          survey        = "acs5",
          state         = "NY",
          keep_geo_vars = T,
          year          = 2014,
          key           = Sys.getenv('CENSUS_KEY'))

# wrangle puma data

puma_incomes <-
  puma_incomes %>% 
  janitor::clean_names() %>% 
  filter(str_detect(name, "Brooklyn|Queens|Manhattan")) %>% 
  mutate(neighborhood = str_split_fixed(name, "--| PUMA", n = 3)[,2]) %>% 
  select(-name, -variable)





# crosswalk is a file with puma income mapped to the SE neighborhood
crosswalk_path <- here("income_map.csv")
crosswalk <- read_csv(crosswalk_path, col_types = "ccic")



# actually choosing buckets
buckets <- c("high-income", "middle-income", "low-income")

crosswalk <- mutate(crosswalk, income_bucket =
             case_when(
      med_income <40000 ~ buckets[3],
      med_income >65000 ~ buckets[1],
      TRUE              ~ buckets[2]))
crosswalk <- mutate(crosswalk, income_bucket = factor(income_bucket, levels = buckets))

# join SE data with PUMA income
df <- inner_join(df, crosswalk)



# get clipped geographies

library(nycgeo)
library(sf)

nyc_geog <- nyc_boundaries(
  geography = "puma",
  filter_by = "borough",
  region = c("brooklyn", "queens", "manhattan")
  )


nyc_geog <- inner_join(nyc_geog, select(crosswalk, geoid, income_bucket), by = "geoid")

bucket_labs <- df %>% 
  count(income_bucket) %>% 
  mutate(lb = str_c(n, " ", income_bucket))




```


As mentioned earlier, this dataset creates a unique opportunity in that is has highly granular neighborhood information. A-priori, neighborhood-level information is clearly relevant to the price of a given rental unit. I mapped *93* Streeteasy Neighborhoods into *39* U.S. Census[^7] *Public Use Microdata Areas*  (PUMA) and then bucketed units into **`r bucket_labs$lb[1]`**, **`r bucket_labs$lb[2]`** and **`r bucket_labs$lb[3]`** `income_buckets`. My approach was to 'eyeball' the distribution of Units by PUMA income and split them as evenly as I could. The key idea here is that the income level of a given neighborhood is correlated with all sorts of variables that cause differences in the price of a rental unit. For instance, low income neighborhoods often have higher crime rates, and, all else equal most people will be willing to pay less money for an apartment in a high-crime neighborhood than for one in a safer neighborhood. 

```{r fig.align="center", out.width="80%", out.height="35%"}
library(patchwork)
library(tidyverse)
library(viridis)
library(scales)

pal <- viridis_pal(option = "magma")(16)[c(13, 9, 5)] 
scale_fill_discrete <- function(...){
  scale_fill_manual(..., values =pal)
}
scale_colour_discrete <- function(...){
  scale_colour_manual(..., values =pal)
}


  income_chlor <-
  nyc_geog %>% 
  ggplot()+
  geom_sf(aes(fill = income_bucket), color = "grey80", lwd = .1)+
  annotate(geom = "text",
           x = Inf,
           y = Inf,
           hjust = 1.3,
           vjust = 2.5,
           label = "Income Buckets of\nPUMA's")+
  theme_void()+
  theme(legend.position = "none")



bucket_distribution_plot <- 
  
df %>% 
  ggplot()+
  geom_histogram(aes(med_income, fill = income_bucket), bins = 30, color = "grey70")+
  
  scale_y_continuous(expand = c(0,0))+
  scale_x_continuous(labels = scales::unit_format(scale = 1e-3, unit = "k", prefix = "$"))+
  
  labs(subtitle = "Units by PUMA median income",
       x = "PUMA Median Income",
       y = "# of Units")+
  
  theme(legend.position = c(.25, .7),
        legend.title = element_blank(),
        legend.key.height = unit(.15, "cm"),
        legend.key.width = unit(.25, "cm"))


d <- "
AABBBB
AABBBB
##BBBB
##BBBB
"

bucket_distribution_plot+income_chlor+plot_layout(design = d)

```


### Neighborhoods as mediators for age effects

To further explore the data, I use the geom_smooth[^8] function from the ggplot2 package in R to make conditional-means plots for both unit size and building age on monthly rent. With this approach I basically build 3 models in order to explore the data. Moving from left to right I plot the conditional means of *the linear model*, *Generalized Additive Model* (GAM)[^9] and a *GAM with an Interaction term*. I'll introduce the GAM's further in a later post, but essentially they are regression methods that allow for more flexibility in modelling linear relationships. I think what laregly stands out here is the inadequacy of the linear model. 

```{r out.width="190%", fig.height = 3.5, fig.align="center"}


linear_plot <- 
ggplot(df, aes(x = age, y = rent))+
  geom_smooth(formula = y~x)+
    geom_rug( sides = "b", color = "black", alpha  = .2, size = .3)+
  labs(title = "Linear model: log(rent)~age",
       y = "Monthly Rent",
       x = "Building Age")+
  scale_y_log10(labels = scales::number_format(scale = 1e-3, suffix = "k", accuracy = 2))+
  theme(plot.title = element_text(size = 9))

smooth_plot <- 
  ggplot(df, aes(x = age, y = rent))+
  geom_smooth()+
    geom_rug( sides = "b", color = "black", alpha  = .2, size = .3)+
  labs(title = "Smooths model: log(rent)~age",
       y = "",
       x = "Building Age")+
  scale_y_log10(labels = scales::number_format(scale = 1e-3, suffix = "k", accuracy = 1))+
  theme(plot.title = element_text(size = 9))
  


age_rent_plt <-
df %>% 
  ggplot(aes(age, rent, color = income_bucket, fill = income_bucket))+
  geom_smooth(key_glyph = draw_key_rect)+
    geom_rug( sides = "b", color = "black", alpha  = .2, size = .3)+
  scale_y_log10(labels = scales::number_format(scale = 1e-3, suffix = "k", accuracy = 1))+
  labs(x = "\nBuilding Age",
       y = "",
       title = str_wrap("Smooths model: log(rent)~age", width = 35),
       subtitle = "by Neighborhood Income")+
  guides(colour = guide_legend(override.aes = list(alpha = 1)))+
  theme(plot.title = element_text(size = 9),
        plot.subtitle = element_text(size = 6.5),
        legend.position = c(.7,.9),
        legend.title = element_blank(),
        legend.key.height = unit(.15, "cm"),
        legend.key.width = unit(.25, "cm"))


# size_rent_plt+age_rent_plt+plot_layout( guides = "collect")

linear_plot+smooth_plot+age_rent_plt

```


## Next Steps

I have explored the economics literature on *filtering effect* in the housing market, gathered some data, enhanced this data with U.S. Census Data, and explored the data with by contrasting a *linear approach*, *a non-linear approach* and a *non-linear approach with mediation*. My ultimate goal is to use this data to make causal inferences about the effects of building-age on the price of rent. Therefore, before I build any statistical model's I'll use Directed Acyclic Graphs (DAGs) to explicitly build causal models of this relationship.  






```{r fig.align="center", fig.height=4, include = F}

library(ggrepel)
med_size_tbl <- 
df %>% 
  filter(bedrooms == 2) %>% 
  group_by(income_bucket) %>% 
  summarise(median_size = median(size_sqft))

size_density_plot <-
df %>% 
  filter(bedrooms==2) %>% 
  ggplot(aes(size_sqft, color = income_bucket))+
  
  geom_line( stat = "density", size = 1, alpha=.7, key_glyph = draw_key_path)+
  
  geom_vline(data = med_size_tbl, 
             aes(xintercept = median_size, 
                 color = income_bucket), 
             size = 1, 
             alpha = .7, 
             linetype = "dotdash",
             key_glyph = draw_key_blank)+
  
  geom_label_repel(data = med_size_tbl, 
                  aes(x = median_size, 
                      label = paste0(median_size, " sqft")),
                  y = c(3.5,2,1), 
                  force = 3,
                  box.padding = 3,
                  key_glyph = draw_key_blank,
                  size  =2)+
  labs(title = "Do high income neighborhoods have larger units?",
       subtitle = str_wrap("Unit size densities for 2 bedroom apartments and median unit size", 35),
       x = expression("Unit size" ~ (ft^2)))+
  scale_x_log10(expand = c(0,0))+
  scale_y_continuous(expand = c(0,0))+
  theme(axis.text.y = element_blank(),
        plot.subtitle = element_text(size = 8),
        plot.title = element_text(size = 11))+
  guides(color = guide_legend(override.aes = list(alpha = 1, fill = pal)))

size_density_plot

```


[^1]:McElreath, Richard. Statistical Rethinking a Bayesian Course with Examples in R and Stan. CRC Press, 2020. 


[^2]:https://www.vox.com/future-perfect/2019/1/22/18192774/oxfam-inequality-report-2019-davos-wealth

[^3]: Coulson, N. Edward, and Eric W. Bond. “A Hedonic Approach to Residential Succession.” The Review of Economics and Statistics, vol. 72, no. 3, 1990, p. 433., doi:10.2307/2109351. 

[^4]: Rosenthal, Stuart S. “Are Private Markets and Filtering a Viable Source of Low-Income Housing? Estimates from a ‘Repeat Income’ Model.” American Economic Review, vol. 104, no. 2, 2014, pp. 687–706., doi:10.1257/aer.104.2.687. 

[^6]: *Housing consumption* is a function of wealth. Wealth distributions approximately follow power laws because wealth creates wealth. 
[^7]: thank you, [{tidycensus}](https://walker-data.com/tidycensus/), a wrapper in R for the Census API
[^8]: https://ggplot2.tidyverse.org/reference/geom_smooth.html
[^9]: Simon Wood, [factor smooths](https://astrostatistics.psu.edu/su07/R/library/mgcv/html/gam.models.html)



