---
title: "NYC Demographics part two: in-migration and out-migration"
author: ''
date: '2020-03-07'
output:
    bookdown::html_document2:
      number_sections: TRUE
      fig_caption: TRUE
slug: in-migration-and-out-migration
categories: []
tags: []
---
```{r setup, echo=F}

knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)


```

```{r}
if(!require(pacman)) install.packages("pacman")
pacman::p_load(
                RPostgres,
                DBI,
                tidyverse,
                here,
                scales,
                faRley,
                ggrepel,
                forcats,
                RColorBrewer,
                grattan,
                egg)

# already-processed ipums microdata
load(file = here("mig_data.RData"))

```
### Brooklyn and Queens see distinct increase in out-migration, starting in 2010

```{r fig.width=4, fig.align='center'}

# aggregation of borough-level out-migration,
out_boro <- 
out_mig %>% 
  group_by(year, borough) %>% 
  summarise(count = sum(per_wt)) %>% 
  mutate(label = 'Out-Migration')
# in-migration
in_boro <- 
  in_mig %>% 
  group_by(year, borough) %>% 
  summarise(count = sum(per_wt)) %>% 
  mutate(label = 'In-Migration')

# combine the two
boro_mig <- 
  out_boro %>% 
  bind_rows(in_boro)

# specific points to hi-light
points <- 
  out_boro %>% 
  filter(borough %in% c('Brooklyn', 'Queens') & year %in% c(2010, 2018)) %>% 
  mutate(labels = 
           paste0(
             as.character(
               round(count/1000, digits = 1)), 'k'))

yr <- c(2006, 2010, 2014, 2018)
yr2 <- str_sub(as.character(yr), 3, -1)
yr3 <- paste0('`', yr2)
  

# plot migration, facet by borough
boro_mig %>% 
  ggplot(aes(x = year, y = count, color = label))+
  geom_line(lwd = 1, alpha = .85)+
  scale_y_continuous(label = unit_format(unit = "k", scale = 1e-3, sep = ""))+
  geom_point(data = points, color = 'black', alpha = .8)+
  scale_x_continuous(breaks = yr, labels = yr3)+
  geom_text_repel(data = points, 
                  color = 'black',
                  fontface = 'bold',
                  size = 3.5,
                  aes(label = labels))+
  scale_color_brewer(type = "qual", palette = 2)+
  facet_wrap(~borough)+

  labs(title = "Plot 1: Brooklyn and queens see \nincreases in out-migration from 2010-2018",
       caption = "Data: IPUMS")+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        legend.title = element_blank(),
        axis.title = element_blank(),
        legend.position = 'top')

```

Plot 1 shows that the de-populationg trends are largely affected by increases in out-migration, particularly in brooklyn and Queens. Its is interesting that Brooklyn and queens have such similar processes with both in-migration and out-migration compared to bronx and manhattan. <br/><br/>

#### Younger and wealthier New-Yorkers are leaving

- **young people:** the spread is greatest in the 30-39 age cohort. It seems likely that this comes from a lot of families moving to the suburbs. I'll have to look at where people are moving to in order to answer this question.

- **rich people:** I thought people were getting priced out of new york but this data negates that intuition. I really don't understand whats going on here. Maybe the suburbs are becoming cooler and everyone wants to move there...


```{r age_cohort, cache=T}
# my local connection where i keep the ipums data
con <- get_swamp(Sys.getenv('LOCAL_UNAME'), Sys.getenv('LOCAL_PW'))


df <- 
  con %>% 
  tbl("nyc_migration") %>% 
  collect() %>% 
  mutate(age_bin1 = factor(age_bin1))

# the two time periods im interested in
df2 <-
  df %>% mutate(window = case_when(
    year %in% 2016:2018 ~ "2016-2018",
    year %in% 2010:2012 ~ "2010-2012"
  ))


```

```{r}
# aggregate out-migration by 10-year age bins
age_piv1 <- 
df2 %>% 
  filter(!is.na(window) & age>19) %>% 
  group_by(window,  age_bin1, nyc_emigrate) %>% 
  summarise(sum = sum(per_wt)) %>% 
  mutate(prop = sum/sum(sum)) %>% 
  filter(nyc_emigrate)

# age plot
age <- ggplot(age_piv1)+
  geom_line(aes(x = age_bin1, 
                y = prop*100, 
                group = window, 
                color = window), lwd = 1)+
  scale_color_brewer(palette = "Paired")+
  scale_y_continuous(breaks = seq(1,5.5,.5),
                     labels = paste0(seq(1,5.5,.5), "%"))+
  labs(title = "who leaves NYC?",
       subtitle = "age cohort",
       x = "age bin (20: 20-29 yo)",
       y = "percent leaving in a given year")+
  theme_minimal()+
  theme(legend.position = "top",
        legend.title = element_blank(),
        panel.grid.minor = element_blank())

# create weighted quintiles for income
df3 <- 
df2 %>% 
  filter(age>25) %>% 
  mutate(inc_percentile = weighted_ntile(income, per_wt, 10)*10)

# aggregate by quintile
income_piv <- 
df3 %>% 
  filter(!is.na(window)) %>% 
  group_by(window, inc_percentile) %>% 
  summarise(mean = weighted.mean(nyc_emigrate))

  income <- ggplot(income_piv)+
  geom_line(aes(x = inc_percentile, 
                y = mean*100, 
                group = window, 
                color = window), lwd = 1)+
    labs(title= "Emigration rate by Income Decile",
         caption = "Data: IPUMS",
         y = "percent leaving in a given year",
         x = "income percentile")+
  scale_color_brewer(palette = "Paired")+
    
    scale_y_continuous(breaks = seq(2.5,5.5, .5),
                       labels = paste0(seq(2.5,5.5, .5), "%"))+
    scale_x_continuous(breaks = seq(10,100,10),
                       labels = paste0(seq(10,100,10), '%'))+
    theme_minimal()+
    theme(panel.grid.minor = element_blank(),
          legend.position = "top",
          legend.title = element_blank())
  
  ggarrange(age, income, nrow = 1)
  
  

  
```

## a couple of puzzles

This data produces more questions than answers. I'll have to go back to 'the drawing board' to think about what questions I should be asking next. I did learn a couple of things from this data:

- Brooklyn and Queens are similar; Bronx and Manhattan are unique

- People are leaving at higher rates. Particularly the 20-40 y/o cohort. The highest income decile had the greatest increase in leaving-rate.

Next I will look at the housing market and try to understand the dynamics between cost-of-living and migration in NYC.
